---
title: "01 EDA Assignment"
output: html_notebook
---
The file "student data for EDA.csv" contains exam, homework, and project scores for 119
students. 
-Using R perform exploratory data analysis. 
-Calculate summary statistics for the data (min, max, mean, standard deviation). 
-Create histograms, 
-run (line) charts, and 
-scatter plots for each column (exam, homework, and project). 
-Clearly, identify any potential information quality issues in the data and provide the output the identified the issue. 
-Include your R and SAS programs.

## load data
```{r}
df <-read.csv("C:/Users/tyler/OneDrive/Documents/GitHub/work_to_show/Classes Spring 2026/BDA Spring 26/student data for EDA.csv",header=TRUE) # FORWARD SLASH AND HEADER TRUE IF HAVE COLUMN NAMES
View(df)
```

## column names
```{r}
# column names
colnames(df) # "Student"  "Exam"     "Homework" "Project" 
#head of the data
head(df)
```

## summary stats
```{r}

summary(df)
"    Student           Exam          Homework         Project      
 Min.   :  1.0   Min.   :30.82   Min.   :-77.73   Min.   : 29.56  
 1st Qu.: 30.5   1st Qu.:66.14   1st Qu.: 73.03   1st Qu.: 73.34  
 Median : 60.0   Median :79.39   Median : 88.34   Median : 87.18  
 Mean   : 60.0   Mean   :75.65   Mean   : 80.80   Mean   : 82.17  
 3rd Qu.: 89.5   3rd Qu.:90.19   3rd Qu.:100.00   3rd Qu.: 99.17  
 Max.   :119.0   Max.   :99.40   Max.   :100.00   Max.   :100.00"
sd(df$Exam) # 18.20467
sd(df$Homework) # 23.82734
sd(df$Project) # 18.30348

```

## histogram
```{r}
hist(df$Exam)
hist(df$Homework) # error on homework 1 value <0
hist(df$Project) 
```

## run chart
```{r}
#plot with ab line
plot(df$Exam,type="b")
abline(h=mean(df$Exam_grade))
plot(df$Homework,type="b") # Value <0
abline(h=mean(df$Homework_grade))
plot(df$Project,type="b")
abline(h=mean(df$Project_grade))
```
## scatter plot
```{r}
# "Student"  "Exam"     "Homework" "Project" 
#correlation
cor(df$Exam,df$Homework)
plot(df$Exam,df$Homework)
cor(df$Homework,df$Project) # NOTE: Entry error on homework
plot(df$Homework,df$Project)
cor(df$Exam,df$Project)
plot(df$Exam,df$Project)
```



