---
title: "BDA Spring 26 Notes"
output: html_notebook
---
# EDA in R (Full Script)
```{r EDA in R (full script), echo=TRUE, message=TRUE, warning=TRUE, paged.print=TRUE}
project_csv <-read.csv("C:/Users/tyler/OneDrive/Documents/GitHub/work_to_show/Classes Spring 2026/BDA Spring 26/Datasets/project.csv",header=TRUE) # FORWARD SLASH AND HEADER TRUE IF HAVE COLUMN NAMES
# column names
colnames(project_csv) # "days_duration" "project_cost" 
#head of the data
head(project_csv)
#summary of the days_duration column
summary(project_csv$days_duration)
#summary of the project_cost column
summary(project_csv$project_cost)
# mean of columns
mean(project_csv$days_duration)
mean(project_csv$project_cost)
# stand dev of columns
sd(project_csv$days_duration)
sd(project_csv$project_cost)
# finding the regression line 
mean(project_csv$days_duration)+3*sd(project_csv$days_duration)
mean(project_csv$days_duration)-3*sd(project_csv$days_duration)
mean(project_csv$project_cost)+3*sd(project_csv$project_cost)
mean(project_csv$project_cost)-3*sd(project_csv$project_cost)
#histogram
hist(project_csv$days_duration)
hist(project_csv$project_cost)
#plot with ab line
plot(project_csv$days_duration,type="b")
abline(h=mean(project_csv$days_duration))
plot(project_csv$project_cost,type="b")
abline(h=mean(project_csv$project_cost))
#correlation
cor(project_csv$days_duration,project_csv$project_cost)
plot(project_csv$days_duration,project_csv$project_cost)
```

# dataForEDA
```{r dataForEDA, echo=TRUE}
EDA <-read.csv ("C:/Users/tyler/OneDrive/Documents/GitHub/work_to_show/Classes Spring 2026/BDA Spring 26/Datasets/dataForEDA.csv", header=TRUE)
#head of data
head(EDA)
#summary of data
summary(EDA)
#mean of data 
mean(EDA$W)
# stand dev data
sd(EDA$W)
sd(EDA$Y)
sd(EDA$X)
sd(EDA$Z)
#mean of data
mean(EDA$X)
mean(EDA$Y)
mean(EDA$Z)
#correlation of data
cor(EDA)
#histogram 
hist(EDA$W)
hist(EDA$X)
hist(EDA$Y)
hist(EDA$Z)
# line plot of data
plot(EDA$W,type="b")
abline(h=mean(EDA$W))
plot(EDA$X,type="b")
plot(EDA$Y,type="b")
plot(EDA$Z,type="b")
# scatter plot of data
plot(EDA$W,EDA$X)
plot(EDA$W,EDA$Y)
plot(EDA$W,EDA$Z)
plot(EDA$X,EDA$Y)
plot(EDA$X,EDA$Z)
plot(EDA$Y,EDA$Z)
```
# one sample mean
```{r one sample mean, echo=TRUE, message=TRUE, warning=TRUE, paged.print=TRUE}
micro <- read.csv("C:/Users/tyler/OneDrive/Documents/GitHub/work_to_show/Classes Spring 2026/BDA Spring 26/Datasets/lAPC one sample ttest.csv", header = TRUE)
t.test(micro$lAPC,mu=6)# tests the column lAPC against the mean of 6

# If the p-value for the t-test is less than 0.05, conclude that the mean is significantly different from 6:   6.976e-14 = 0.00000000000006976     which is well less than 0.05.  We reject the null hypothesis.

# Reject the null hypothesis (H0: m = 6.0)

# The mean log10 APC level was found to be significantly different from the required quality limit of 6.0 log10APC

# The sample mean was 3.05 log10 APC, we are 95% certain the true mean is between 2.60 and 3.50 log10 APC; the food company can comfortably say the average log10 APC level is well less than the 6.0 requirement

```
# Two Sample t-test unequal
```{r Two Sample t-test, variances assumed unequal – R code, echo=TRUE, message=TRUE, warning=TRUE, paged.print=TRUE}
nfl<-read.csv("C:/Users/tyler/OneDrive/Documents/GitHub/work_to_show/Classes Spring 2026/BDA Spring 26/Datasets/nfl data 2019.csv",header=TRUE)
nfl$Div<-as.factor(nfl$Div)
t.test(OffPtsPerG ~ Div, data = nfl)

# The p-value is 0.169 (greater than 0.05) we fail to reject the null hypothesis.

```

# Two Sample t-test equal
```{r Two Sample t-test, variances assumed equal –R Code, echo=TRUE, message=TRUE, warning=TRUE, paged.print=TRUE}
nfl<-read.csv("C:/Users/tyler/OneDrive/Documents/GitHub/work_to_show/Classes Spring 2026/BDA Spring 26/Datasets/nfl data 2019.csv",header=TRUE)
nfl$Div<-as.factor(nfl$Div) 
t.test(OffPtsPerG ~ Div, data = nfl,var.equal = TRUE)

# The p-value is 0.1686 (greater than 0.05) again, we fail to reject the null hypothesis.


```

# Folded F test
```{r R for Folded F test, echo=TRUE, message=TRUE, warning=TRUE, paged.print=TRUE}
nfl<-read.csv("C:/Users/tyler/OneDrive/Documents/GitHub/work_to_show/Classes Spring 2026/BDA Spring 26/Datasets/nfl data 2019.csv",header=TRUE)
group1 <- nfl$OffPtsPerG[nfl$Div == "AFC"]
group2 <- nfl$OffPtsPerG[nfl$Div == "NFC"]
var.test(group1, group2)

# The p-value is 0.4931 (greater than 0.05), we fail to reject the null hypothesis.

```

# paired t test
```{r R for paired t test, echo=TRUE, message=TRUE, warning=TRUE, paged.print=TRUE}
act<-read.csv("C:/Users/tyler/OneDrive/Documents/GitHub/work_to_show/Classes Spring 2026/BDA Spring 26/Datasets/bda act data.csv",header=TRUE)
t.test(act$ACT_Math, act$ACT_English, paired = TRUE)


#The p-value is 0.004438 (less than 0.05), we reject the null hypothesis.

```

# Power
```{r Power, echo=TRUE, message=TRUE, warning=TRUE, paged.print=TRUE}

power.t.test(n = 30, # sample size in each group      
             delta = .7, #effect size (difference in means that needs to be detected) 
             sd = 0.9, #pooled standard deviation
             sig.level = 0.05, #level of significance for the hypothesis test
             type = "two.sample", #type of t-test
             alternative = "two.sided") #number of tails

```

# Anova
```{r Anova, echo=TRUE, message=TRUE, warning=TRUE, paged.print=TRUE}
apc<-read.csv("C:/Users/tyler/OneDrive/Documents/GitHub/work_to_show/Classes Spring 2026/BDA Spring 26/Datasets/new micro treatment.csv",header=TRUE)
head(apc)
colnames(apc) # "id"       "trt"      "log10APC"
# trt column as factor
# $word <- as.factor( $word)
apc$trt<-as.factor(apc$trt)
# start of anova 
# fit<-aov(col_to_mean ~ unique_identifier, data=data)
fit<-aov(log10APC~trt,data=apc)
summary(fit)
# tukey test for mult comparisons of means 
TukeyHSD(fit,"trt")
# grouped summaries
# aggregate(col_to_mean ~ unique_identifier, data = data, FUN = function)
aggregate(log10APC~trt, data = apc, FUN = mean)
library(agricolae) # to get duncan test
# compare all treatment means
# dun<-duncan.test(fit, "unique_identifier", group = TRUE)
dun<-duncan.test(fit, "trt", group = TRUE)
print(dun)

# Example: 
SOAP<-read.csv("C:/Users/tyler/OneDrive/Documents/GitHub/work_to_show/Classes Spring 2026/BDA Spring 26/Datasets/soap.csv",header=TRUE)
head(SOAP)
colnames(SOAP) # "id"     "Brand"  "Rating"
# Brand column as factor
SOAP$Brand<-as.factor(SOAP$Brand)
# start of anova 
fit<-aov(Rating~Brand,data=SOAP)
summary(fit)
# tukey test for mult comparisons of means 
TukeyHSD(fit,"Brand")
# grouped summaries
aggregate(Rating~Brand, data = SOAP, FUN = mean)
library(agricolae) # to get duncan test
# compare all treatment means
dun<-duncan.test(fit, "Rating", group = TRUE)
print(dun)
```

# Block Design
```{r}
pt<-read.csv("C:/Users/tyler/OneDrive/Documents/GitHub/work_to_show/Classes Spring 2026/BDA Spring 26/Datasets/pt data for blocking.csv",header=TRUE)
head(pt)
pt$store <- as.factor(pt$store)
pt$brand <- as.factor(pt$brand)
#install.packages("agricolae") # if not installed
library(agricolae)
fit <- aov(sales ~ brand + store, data = pt)
summary(fit)
duncan_results <- duncan.test(fit, "brand", group = TRUE)
print(duncan_results)
```
# Scenario
```{r}
s1<-read.csv("C:/Users/tyler/OneDrive/Documents/GitHub/work_to_show/Classes Spring 2026/BDA Spring 26/Datasets/s1 full factorial.csv",header=TRUE)
head(s1)
s1$humidity <- as.factor(s1$humidity)
s1$airflow <- as.factor(s1$airflow)
fit <- aov(yield ~ humidity * airflow, data = s1)
summary(fit)

s2<-read.csv("C:/Users/tyler/OneDrive/Documents/GitHub/work_to_show/Classes Spring 2026/BDA Spring 26/Datasets/s2 full factorial.csv",header=TRUE)
head(s2)
s2$humidity <- as.factor(s2$humidity)
s2$airflow <- as.factor(s2$airflow)
fit <- aov(yield ~ humidity * airflow, data = s2)
summary(fit)
duncan_results <- duncan.test(fit, "humidity", group = TRUE)
print(duncan_results)

s3<-read.csv("C:/Users/tyler/OneDrive/Documents/GitHub/work_to_show/Classes Spring 2026/BDA Spring 26/Datasets/s3 full factorial.csv",header=TRUE)
head(s3)
s3$humidity <- as.factor(s3$humidity)
s3$airflow <- as.factor(s3$airflow)
fit <- aov(yield ~ humidity * airflow, data = s3)
summary(fit)

library(agricolae)
fit <- aov(yield ~ humidity, data = s3)
summary(fit)
duncan_results <- duncan.test(fit, "humidity", group = TRUE)
print(duncan_results)
fit <- aov(yield ~ airflow, data = s3)
summary(fit)
duncan_results <- duncan.test(fit, "airflow", group = TRUE)
print(duncan_results)

s4<-read.csv("C:/Users/tyler/OneDrive/Documents/GitHub/work_to_show/Classes Spring 2026/BDA Spring 26/Datasets/s4 full factorial.csv",header=TRUE)
head(s4)
s4$humidity <- as.factor(s4$humidity)
s4$airflow <- as.factor(s4$airflow)
fit <- aov(yield ~ humidity * airflow, data = s4)
summary(fit)
s4$ha <- paste(s4$humidity, s4$airflow, sep = " ")
head(s4)
#install.packages("agricolae") # if not installed
# library(agricolae)
fit <- aov(yield ~ ha, data = s4)
summary(fit)
duncan_results <- duncan.test(fit, "ha", group = TRUE)
print(duncan_results)

s4$ha <- paste(s4$humidity, s4$airflow, sep = " ")
head(s4)
#install.packages("agricolae") # if not installed
library(agricolae)
fit <- aov(yield ~ ha, data = s4)
summary(fit)
duncan_results <- duncan.test(fit, "ha", group = TRUE)
print(duncan_results)
```
# Covariate Example
```{r}
# Covariate Example
dbp<-read.csv("C:/Users/tyler/OneDrive/Documents/GitHub/work_to_show/Classes Spring 2026/BDA Spring 26/Datasets/dbp data for covariate.csv",header=TRUE)
head(dbp)
dbp$Treatment <- as.factor(dbp$Treatment)
fit <- lm(DBP ~ Treatment + Weight, data = dbp)
anova(fit)

## Covariate follow-up
#install.packages("emmeans") # run once, if not installed
library(emmeans)
emm <- emmeans(fit, ~ Treatment)
pairs(emm)
emm
```
<<<<<<< Updated upstream
# Regression
```{r}
# Regression
house <-read.csv("C:/Users/tyler/OneDrive/Documents/GitHub/work_to_show/Classes Spring 2026/BDA Spring 26/Datasets/housing.csv", header=TRUE)
=======

# DOE Assignment
```{r}
circle <- read.csv("C:/Users/tyler/OneDrive/Documents/GitHub/work_to_show/Classes Spring 2026/BDA Spring 26/Datasets/circle.csv")
fit<-aov(accidents ~ type + Month,data = circle) # block design
summary(fit) # primary factor = type, block = month, are significant 
# install.packages("agricolae")
library(agricolae)
duncan_results<-duncan.test(fit,"type",group = TRUE) # duncan mean seperation 
print(duncan_results)
# Conclusion: The addition of traffic circles has reduced the number of accidents that have occurred:
# Type of Intersection: Mean of accidents
# Circle 6.583a
# Light 14.167b
# ab means with different letters were found to be significantly different at the 0.05 level.

# Block Design
library(agricolae)
clothing <- read.csv("C:/Users/tyler/OneDrive/Documents/GitHub/work_to_show/Classes Spring 2026/BDA Spring 26/Datasets/clothing.csv")
fit<-aov(sales ~ sale*product,data = clothing) 
summary(fit) # sale a product are both signficiant & impacts the sale <2e-16. The main effects of products and sales are also significant
clothing$ps<-paste(clothing$product,clothing$sale,sep = "") # CREATE PRODUT AND SALES COLUMN  
fit<-aov(sales~ps,data = clothing) # ANOTHER ANOVA OF SALES ~ PRODUCT SALES COLUMN 
duncan_results<-duncan.test(fit,"ps", group = TRUE) # do this because of significance. Used to see how 4 means differed. 
print(duncan_results)
# conclusion:  Shirts on sale had the highest mean sales, followed by shirts not on sales. Sales of pants were significantly
# lower than sales of shirts. However, “not on sale” or “on sale” could not be shown to impact pants sales. The
# differences may be summarized in the visual below:
# Sale No Sale
# Shirt $9,801 a $7,760 b
# Pants $5,871 c $5,753 c
# ab means with different letters were found to be significantly different at the 0.05 level.

online <- read.csv("C:/Users/tyler/OneDrive/Documents/GitHub/work_to_show/Classes Spring 2026/BDA Spring 26/Datasets/online.csv")
fit<- lm(FinalGrade~Format+GPA,data=online) # COVARIATE MODEL
anova(fit) # ANOVA OF FIT gpa and format both sig
# install.packages("emmeans")
library(emmeans) 
emm <- emmeans(fit, ~ Format)
pairs(emm) # p-value is significant both covariate and format impact the final grade
emm 
# Conclusion: The impact of the format of class (online or in a classroom) was found to be significant. The in-person
# method of instruction significantly increased average final grade (by 9 points on average). The differences can be
# summarized as follows:
# Format Final Grade Mean
# Classroom 79.2 a
# Online 70.3 b
# ab means with different letters were found to be significantly different at the 0.05 level.
```
# REGRESSION 
```{r}
library(tidyverse)
house <- read.csv("C:/Users/tyler/OneDrive/Documents/GitHub/work_to_show/Classes Spring 2026/BDA Spring 26/Datasets/housing.csv")
colnames(house)
>>>>>>> Stashed changes
head(house)
model <- lm(saleprice ~ sqft, data = house)
summary(model)
#below this line “extra” code to provide a visual plot
plot(house$sqft,house$saleprice)
abline(model)
pred_interval <- predict(model, newdata=house, interval="prediction",level = 0.95)
lines(house$sqft, pred_interval[,2])
lines(house$sqft, pred_interval[,3])
<<<<<<< Updated upstream
```

```{r}
# Multi Regression
house2 <-read.csv("C:/Users/tyler/OneDrive/Documents/GitHub/work_to_show/Classes Spring 2026/BDA Spring 26/Datasets/housing2.csv", header=TRUE)
head(house2)
model <- lm(SalePrice ~ SqFt+Age+Location+Lotsize+Condition , data = house2)
#alternate syntax to include all other columns:
#model <- lm(SalePrice ~ ., data = house2)
summary(model)
```
=======


```

```{r}
house2 <- read.csv('C:/Users/tyler/OneDrive/Documents/GitHub/work_to_show/Classes Spring 2026/BDA Spring 26/Datasets/housing2.csv', sep = ',')
head(house2)
model <- lm(SalePrice ~ SqFt+Age+Location+Lotsize+Condition , data = house2)
#alternate syntax to include all other columns:
# model <- lm(SalePrice ~ ., data = house2)
summary(model)
>>>>>>> Stashed changes

# Categroical Analysis
R Code for 1-Sample p-test
```{r}
prop.test(39, 60, p = 0.5, correct=FALSE)
```
R code for 2-Proportion Test (Contingency Table and Odds Ratio)
```{r}
# Create data frame
p <- data.frame(
  method = c("A", "A", "B", "B"),
  quality = c("Good", "Defective", "Good", "Defective"),
  count = c(75, 25, 87, 13)
)

# Create contingency table (apply weights)
tab <- xtabs(count ~ method + quality, data = p)
tab

# Chi-square test of independence
cs <- chisq.test(tab)
cs

# Expected cell counts
cs$expected

# Fisher's Exact Test
fisher.test(tab)

# Odds Ratio (Wald method)
library(epitools)
oddsratio(tab, method = "wald")
```
McNemar’s Test for Matched Pairs Proportions
```{r}
pm <- data.frame(
before = c("PreMed", "PreMed", "NotPM", "NotPM"),
after = c("PreMed", "NotPM", "PreMed", "NotPM"),
count = c(29, 21, 6, 34)
)
# Turn into a contingency table with weights
tab <- xtabs(count ~ before + after, data = pm)
tab
# Agreement statistics
mcnemar.test(tab,correct=FALSE)
```
Assignment
```{r}
prop.test(71, 100, p = 0.5, correct=FALSE)
```

```{r}
# Create data frame
p <- data.frame(
  firstgen = c("NO", "NO", "YES", "YES"),
  OUTCOME = c("DID NOT FINISH", "COMPLETE", "DID NOT FINISH", "COMPLETE"),
  count = c(21, 19, 22, 15)
)

# Create contingency table (apply weights)
tab <- xtabs(count ~ firstgen + OUTCOME, data = p)
tab

# Chi-square test of independence
cs <- chisq.test(tab)
cs

# Expected cell counts
cs$expected

# Fisher's Exact Test
fisher.test(tab)

# Odds Ratio (Wald method)
library(epitools)
oddsratio(tab, method = "wald")
```

```{r}
pm <- data.frame(
before = c("A", "A", "B", "B"),
after = c("A", "B", "A", "B"),
count = c(71, 31, 14, 84)
)
# Turn into a contingency table with weights
tab <- xtabs(count ~ before + after, data = pm)
tab
# Agreement statistics
mcnemar.test(tab,correct=FALSE)
```

```{r}
nuggets <- read.csv('C:/Users/tyler/OneDrive/Documents/GitHub/work_to_show/Classes Spring 2026/BDA Spring 26/Datasets/nuggets.csv', sep = ',')
nuggets

head(nuggets)
model <- lm(temp ~ weight, data = nuggets)
summary(model)
#below this line “extra” code to provide a visual plot
plot(nuggets$weight,nuggets$temp)
abline(model)
pred_interval <- predict(model, newdata=nuggets, interval="prediction",level = 0.95)
lines(nuggets$weight, pred_interval[,2])
lines(nuggets$weight, pred_interval[,3])

```

```{r}
tendersmr <- read.csv('C:/Users/tyler/OneDrive/Documents/GitHub/work_to_show/Classes Spring 2026/BDA Spring 26/Datasets/tendersmr.csv', sep = ',')

head(tendersmr)
model <- lm(farmscore ~ Gender+live+tenders , data = tendersmr)
#alternate syntax to include all other columns:
# model <- lm(farmscore ~ ., data = tendersmr)
summary(model)
```

# P-VALUE REJECTION
```{r P-VALUE REJECTION, echo=TRUE, message=TRUE, warning=TRUE, paged.print=TRUE}
p_val <- 2e-16
level_sig <- 0.05

if (p_val <= level_sig) {
  print("Reject")
} else {
  print("Fail to reject")
}
```

