---
title: "BDA Spring 26 Notes"
output: html_notebook
---
# EDA in R (Full Script)
```{r EDA in R (full script), echo=TRUE, message=TRUE, warning=TRUE, paged.print=TRUE}
project_csv <-read.csv("C:/Users/tyler/OneDrive/Documents/GitHub/work_to_show/Classes Spring 2026/BDA Spring 26/Datasets/project.csv",header=TRUE) # FORWARD SLASH AND HEADER TRUE IF HAVE COLUMN NAMES
# column names
colnames(project_csv) # "days_duration" "project_cost" 
#head of the data
head(project_csv)
#summary of the days_duration column
summary(project_csv$days_duration)
#summary of the project_cost column
summary(project_csv$project_cost)
# mean of columns
mean(project_csv$days_duration)
mean(project_csv$project_cost)
# stand dev of columns
sd(project_csv$days_duration)
sd(project_csv$project_cost)
# finding the regression line 
mean(project_csv$days_duration)+3*sd(project_csv$days_duration)
mean(project_csv$days_duration)-3*sd(project_csv$days_duration)
mean(project_csv$project_cost)+3*sd(project_csv$project_cost)
mean(project_csv$project_cost)-3*sd(project_csv$project_cost)
#histogram
hist(project_csv$days_duration)
hist(project_csv$project_cost)
#plot with ab line
plot(project_csv$days_duration,type="b")
abline(h=mean(project_csv$days_duration))
plot(project_csv$project_cost,type="b")
abline(h=mean(project_csv$project_cost))
#correlation
cor(project_csv$days_duration,project_csv$project_cost)
plot(project_csv$days_duration,project_csv$project_cost)
```

# dataForEDA
```{r dataForEDA, echo=TRUE}
EDA <-read.csv ("C:/Users/tyler/OneDrive/Documents/GitHub/work_to_show/Classes Spring 2026/BDA Spring 26/Datasets/dataForEDA.csv", header=TRUE)
#head of data
head(EDA)
#summary of data
summary(EDA)
#mean of data 
mean(EDA$W)
# stand dev data
sd(EDA$W)
sd(EDA$Y)
sd(EDA$X)
sd(EDA$Z)
#mean of data
mean(EDA$X)
mean(EDA$Y)
mean(EDA$Z)
#correlation of data
cor(EDA)
#histogram 
hist(EDA$W)
hist(EDA$X)
hist(EDA$Y)
hist(EDA$Z)
# line plot of data
plot(EDA$W,type="b")
abline(h=mean(EDA$W))
plot(EDA$X,type="b")
plot(EDA$Y,type="b")
plot(EDA$Z,type="b")
# scatter plot of data
plot(EDA$W,EDA$X)
plot(EDA$W,EDA$Y)
plot(EDA$W,EDA$Z)
plot(EDA$X,EDA$Y)
plot(EDA$X,EDA$Z)
plot(EDA$Y,EDA$Z)
```
# one sample mean
```{r one sample mean, echo=TRUE, message=TRUE, warning=TRUE, paged.print=TRUE}
micro <- read.csv("C:/Users/tyler/OneDrive/Documents/GitHub/work_to_show/Classes Spring 2026/BDA Spring 26/Datasets/lAPC one sample ttest.csv", header = TRUE)
t.test(micro$lAPC,mu=6)# tests the column lAPC against the mean of 6

# If the p-value for the t-test is less than 0.05, conclude that the mean is significantly different from 6:   6.976e-14 = 0.00000000000006976     which is well less than 0.05.  We reject the null hypothesis.

# Reject the null hypothesis (H0: m = 6.0)

# The mean log10 APC level was found to be significantly different from the required quality limit of 6.0 log10APC

# The sample mean was 3.05 log10 APC, we are 95% certain the true mean is between 2.60 and 3.50 log10 APC; the food company can comfortably say the average log10 APC level is well less than the 6.0 requirement

```
# Two Sample t-test unequal
```{r Two Sample t-test, variances assumed unequal – R code, echo=TRUE, message=TRUE, warning=TRUE, paged.print=TRUE}
nfl<-read.csv("C:/Users/tyler/OneDrive/Documents/GitHub/work_to_show/Classes Spring 2026/BDA Spring 26/Datasets/nfl data 2019.csv",header=TRUE)
nfl$Div<-as.factor(nfl$Div)
t.test(OffPtsPerG ~ Div, data = nfl)

# The p-value is 0.169 (greater than 0.05) we fail to reject the null hypothesis.

```

# Two Sample t-test equal
```{r Two Sample t-test, variances assumed equal –R Code, echo=TRUE, message=TRUE, warning=TRUE, paged.print=TRUE}
nfl<-read.csv("C:/Users/tyler/OneDrive/Documents/GitHub/work_to_show/Classes Spring 2026/BDA Spring 26/Datasets/nfl data 2019.csv",header=TRUE)
nfl$Div<-as.factor(nfl$Div) 
t.test(OffPtsPerG ~ Div, data = nfl,var.equal = TRUE)

# The p-value is 0.1686 (greater than 0.05) again, we fail to reject the null hypothesis.


```

# Folded F test
```{r R for Folded F test, echo=TRUE, message=TRUE, warning=TRUE, paged.print=TRUE}
nfl<-read.csv("C:/Users/tyler/OneDrive/Documents/GitHub/work_to_show/Classes Spring 2026/BDA Spring 26/Datasets/nfl data 2019.csv",header=TRUE)
group1 <- nfl$OffPtsPerG[nfl$Div == "AFC"]
group2 <- nfl$OffPtsPerG[nfl$Div == "NFC"]
var.test(group1, group2)

# The p-value is 0.4931 (greater than 0.05), we fail to reject the null hypothesis.

```

# paired t test
```{r R for paired t test, echo=TRUE, message=TRUE, warning=TRUE, paged.print=TRUE}
act<-read.csv("C:/Users/tyler/OneDrive/Documents/GitHub/work_to_show/Classes Spring 2026/BDA Spring 26/Datasets/bda act data.csv",header=TRUE)
t.test(act$ACT_Math, act$ACT_English, paired = TRUE)


#The p-value is 0.004438 (less than 0.05), we reject the null hypothesis.

```

# Power
```{r Power, echo=TRUE, message=TRUE, warning=TRUE, paged.print=TRUE}

power.t.test(n = 30, # sample size in each group      
             delta = .7, #effect size (difference in means that needs to be detected) 
             sd = 0.9, #pooled standard deviation
             sig.level = 0.05, #level of significance for the hypothesis test
             type = "two.sample", #type of t-test
             alternative = "two.sided") #number of tails

```

# Anova
```{r Anova, echo=TRUE, message=TRUE, warning=TRUE, paged.print=TRUE}
apc<-read.csv("C:/Users/tyler/OneDrive/Documents/GitHub/work_to_show/Classes Spring 2026/BDA Spring 26/Datasets/new micro treatment.csv",header=TRUE)
head(apc)
colnames(apc) # "id"       "trt"      "log10APC"
# trt column as factor
# $word <- as.factor( $word)
apc$trt<-as.factor(apc$trt)
# start of anova 
# fit<-aov(col_to_mean ~ unique_identifier, data=data)
fit<-aov(log10APC~trt,data=apc)
summary(fit)
# tukey test for mult comparisons of means 
TukeyHSD(fit,"trt")
# grouped summaries
# aggregate(col_to_mean ~ unique_identifier, data = data, FUN = function)
aggregate(log10APC~trt, data = apc, FUN = mean)
library(agricolae) # to get duncan test
# compare all treatment means
# dun<-duncan.test(fit, "unique_identifier", group = TRUE)
dun<-duncan.test(fit, "trt", group = TRUE)
print(dun)

# Example: 
SOAP<-read.csv("C:/Users/tyler/OneDrive/Documents/GitHub/work_to_show/Classes Spring 2026/BDA Spring 26/Datasets/soap.csv",header=TRUE)
head(SOAP)
colnames(SOAP) # "id"     "Brand"  "Rating"
# Brand column as factor
SOAP$Brand<-as.factor(SOAP$Brand)
# start of anova 
fit<-aov(Rating~Brand,data=SOAP)
summary(fit)
# tukey test for mult comparisons of means 
TukeyHSD(fit,"Brand")
# grouped summaries
aggregate(Rating~Brand, data = SOAP, FUN = mean)
library(agricolae) # to get duncan test
# compare all treatment means
dun<-duncan.test(fit, "Rating", group = TRUE)
print(dun)
```

# Block Design
```{r}
pt<-read.csv("C:/Users/tyler/OneDrive/Documents/GitHub/work_to_show/Classes Spring 2026/BDA Spring 26/Datasets/pt data for blocking.csv",header=TRUE)
head(pt)
pt$store <- as.factor(pt$store)
pt$brand <- as.factor(pt$brand)
#install.packages("agricolae") # if not installed
library(agricolae)
fit <- aov(sales ~ brand + store, data = pt)
summary(fit)
duncan_results <- duncan.test(fit, "brand", group = TRUE)
print(duncan_results)
```
# 2x2 Factorial
```{r}
s1<-read.csv("C:/Users/tyler/OneDrive/Documents/GitHub/work_to_show/Classes Spring 2026/BDA Spring 26/Datasets/s1 full factorial.csv",header=TRUE)
head(s1)
s1$humidity <- as.factor(s1$humidity)
s1$airflow <- as.factor(s1$airflow)
fit <- aov(yield ~ humidity * airflow, data = s1)
summary(fit)

s2<-read.csv("C:/Users/tyler/OneDrive/Documents/GitHub/work_to_show/Classes Spring 2026/BDA Spring 26/Datasets/s2 full factorial.csv",header=TRUE)
head(s2)
s2$humidity <- as.factor(s2$humidity)
s2$airflow <- as.factor(s2$airflow)
fit <- aov(yield ~ humidity * airflow, data = s2)
summary(fit)
duncan_results <- duncan.test(fit, "humidity", group = TRUE)
print(duncan_results)

s3<-read.csv("C:/Users/tyler/OneDrive/Documents/GitHub/work_to_show/Classes Spring 2026/BDA Spring 26/Datasets/s3 full factorial.csv",header=TRUE)
head(s3)
s3$humidity <- as.factor(s3$humidity)
s3$airflow <- as.factor(s3$airflow)
fit <- aov(yield ~ humidity * airflow, data = s3)
summary(fit)

library(agricolae)
fit <- aov(yield ~ humidity, data = s3)
summary(fit)
duncan_results <- duncan.test(fit, "humidity", group = TRUE)
print(duncan_results)
fit <- aov(yield ~ airflow, data = s3)
summary(fit)
duncan_results <- duncan.test(fit, "airflow", group = TRUE)
print(duncan_results)

s4<-read.csv("C:/Users/tyler/OneDrive/Documents/GitHub/work_to_show/Classes Spring 2026/BDA Spring 26/Datasets/s4 full factorial.csv",header=TRUE)
head(s4)
s4$humidity <- as.factor(s4$humidity)
s4$airflow <- as.factor(s4$airflow)
fit <- aov(yield ~ humidity * airflow, data = s4)
summary(fit)
s4$ha <- paste(s4$humidity, s4$airflow, sep = " ")
head(s4)
#install.packages("agricolae") # if not installed
# library(agricolae)
fit <- aov(yield ~ ha, data = s4)
summary(fit)
duncan_results <- duncan.test(fit, "ha", group = TRUE)
print(duncan_results)

s4$ha <- paste(s4$humidity, s4$airflow, sep = " ")
head(s4)
#install.packages("agricolae") # if not installed
library(agricolae)
fit <- aov(yield ~ ha, data = s4)
summary(fit)
duncan_results <- duncan.test(fit, "ha", group = TRUE)
print(duncan_results)
```
# Covariate Example
```{r}
# Covariate Example
dbp<-read.csv("C:/Users/tyler/OneDrive/Documents/GitHub/work_to_show/Classes Spring 2026/BDA Spring 26/Datasets/dbp data for covariate.csv",header=TRUE)
head(dbp)
dbp$Treatment <- as.factor(dbp$Treatment) # treatment as factors
# fit <- lm(col_find ~ factor1 + factor2, data = df)
fit <- lm(DBP ~ Treatment + Weight, data = dbp)
anova(fit) # anova table

## Covariate follow-up
#install.packages("emmeans") # run once, if not installed
library(emmeans)
emm <- emmeans(fit, ~ Treatment)
pairs(emm)
emm
```

# DOE Assignment
```{r}
circle <- read.csv("C:/Users/tyler/OneDrive/Documents/GitHub/work_to_show/Classes Spring 2026/BDA Spring 26/Datasets/circle.csv")
fit<-aov(accidents ~ type + Month,data = circle) # block design
summary(fit) # primary factor = type, block = month, are significant 
# install.packages("agricolae")
library(agricolae)
duncan_results<-duncan.test(fit,"type",group = TRUE) # duncan mean seperation 
print(duncan_results)
# Conclusion: The addition of traffic circles has reduced the number of accidents that have occurred:
# Type of Intersection: Mean of accidents
# Circle 6.583a
# Light 14.167b
# ab means with different letters were found to be significantly different at the 0.05 level.

# Block Design
library(agricolae)
clothing <- read.csv("C:/Users/tyler/OneDrive/Documents/GitHub/work_to_show/Classes Spring 2026/BDA Spring 26/Datasets/clothing.csv")
fit<-aov(sales ~ sale*product,data = clothing) 
summary(fit) # sale a product are both signficiant & impacts the sale <2e-16. The main effects of products and sales are also significant
clothing$ps<-paste(clothing$product,clothing$sale,sep = "") # CREATE PRODUT AND SALES COLUMN  
fit<-aov(sales~ps,data = clothing) # ANOTHER ANOVA OF SALES ~ PRODUCT SALES COLUMN 
duncan_results<-duncan.test(fit,"ps", group = TRUE) # do this because of significance. Used to see how 4 means differed. 
print(duncan_results)
# conclusion:  Shirts on sale had the highest mean sales, followed by shirts not on sales. Sales of pants were significantly
# lower than sales of shirts. However, “not on sale” or “on sale” could not be shown to impact pants sales. The
# differences may be summarized in the visual below:
# Sale No Sale
# Shirt $9,801 a $7,760 b
# Pants $5,871 c $5,753 c
# ab means with different letters were found to be significantly different at the 0.05 level.

online <- read.csv("C:/Users/tyler/OneDrive/Documents/GitHub/work_to_show/Classes Spring 2026/BDA Spring 26/Datasets/online.csv")
fit<- lm(FinalGrade~Format+GPA,data=online) # COVARIATE MODEL
anova(fit) # ANOVA OF FIT gpa and format both sig
# install.packages("emmeans")
library(emmeans) 
emm <- emmeans(fit, ~ Format)
pairs(emm) # p-value is significant both covariate and format impact the final grade
emm 
# Conclusion: The impact of the format of class (online or in a classroom) was found to be significant. The in-person
# method of instruction significantly increased average final grade (by 9 points on average). The differences can be
# summarized as follows:
# Format Final Grade Mean
# Classroom 79.2 a
# Online 70.3 b
# ab means with different letters were found to be significantly different at the 0.05 level.
```
# REGRESSION 
```{r}
library(tidyverse)
house <- read.csv("C:/Users/tyler/OneDrive/Documents/GitHub/work_to_show/Classes Spring 2026/BDA Spring 26/Datasets/housing.csv")
colnames(house)

head(house)
model <- lm(saleprice ~ sqft, data = house)
summary(model)
#below this line “extra” code to provide a visual plot
plot(house$sqft,house$saleprice)
abline(model)
pred_interval <- predict(model, newdata=house, interval="prediction",level = 0.95)
lines(house$sqft, pred_interval[,2])
lines(house$sqft, pred_interval[,3])

```

```{r}
# Multi Regression
house2 <-read.csv("C:/Users/tyler/OneDrive/Documents/GitHub/work_to_show/Classes Spring 2026/BDA Spring 26/Datasets/housing2.csv", header=TRUE)
head(house2)
model <- lm(SalePrice ~ SqFt+Age+Location+Lotsize+Condition , data = house2)
#alternate syntax to include all other columns:
#model <- lm(SalePrice ~ ., data = house2)
summary(model)

house2 <- read.csv('C:/Users/tyler/OneDrive/Documents/GitHub/work_to_show/Classes Spring 2026/BDA Spring 26/Datasets/housing2.csv', sep = ',')
head(house2)
model <- lm(SalePrice ~ SqFt+Age+Location+Lotsize+Condition , data = house2)
#alternate syntax to include all other columns:
# model <- lm(SalePrice ~ ., data = house2)
summary(model)
```

```{r}
house2 <- read.csv('C:/Users/tyler/OneDrive/Documents/GitHub/work_to_show/Classes Spring 2026/BDA Spring 26/Datasets/housing2.csv', sep = ',')
head(house2)
model <- lm(SalePrice ~ SqFt+Age+Location+Lotsize+Condition , data = house2)
#alternate syntax to include all other columns:
# model <- lm(SalePrice ~ ., data = house2)
summary(model)
```

```{r}
house2 <- read.csv('C:/Users/tyler/OneDrive/Documents/GitHub/work_to_show/Classes Spring 2026/BDA Spring 26/Datasets/housing2.csv', sep = ',')
head(house2)
model <- lm(SalePrice ~ SqFt+Age+Location+Lotsize+Condition , data = house2)
#alternate syntax to include all other columns:
# model <- lm(SalePrice ~ ., data = house2)
summary(model)


# Categroical Analysis
prop.test(39, 60, p = 0.5, correct=FALSE)
```
# R code for 2-Proportion Test (Contingency Table and Odds Ratio)
```{r}
# Create data frame
p <- data.frame(
  method = c("A", "A", "B", "B"),
  quality = c("Good", "Defective", "Good", "Defective"),
  count = c(75, 25, 87, 13)
)

# Create contingency table (apply weights)
tab <- xtabs(count ~ method + quality, data = p)
tab

# Chi-square test of independence
cs <- chisq.test(tab)
cs

# Expected cell counts
cs$expected

# Fisher's Exact Test
fisher.test(tab)

# Odds Ratio (Wald method)
library(epitools)
oddsratio(tab, method = "wald")
```
# McNemar’s Test for Matched Pairs Proportions
```{r}
pm <- data.frame(
before = c("PreMed", "PreMed", "NotPM", "NotPM"),
after = c("PreMed", "NotPM", "PreMed", "NotPM"),
count = c(29, 21, 6, 34)
)
# Turn into a contingency table with weights
tab <- xtabs(count ~ before + after, data = pm)
tab
# Agreement statistics
mcnemar.test(tab,correct=FALSE)
```
# Assignment
```{r}
prop.test(71, 100, p = 0.5, correct=FALSE)
```

```{r}
# Create data frame
p <- data.frame(
  firstgen = c("NO", "NO", "YES", "YES"),
  OUTCOME = c("DID NOT FINISH", "COMPLETE", "DID NOT FINISH", "COMPLETE"),
  count = c(21, 19, 22, 15)
)

# Create contingency table (apply weights)
tab <- xtabs(count ~ firstgen + OUTCOME, data = p)
tab

# Chi-square test of independence
cs <- chisq.test(tab)
cs

# Expected cell counts
cs$expected

# Fisher's Exact Test
fisher.test(tab)

# Odds Ratio (Wald method)
library(epitools)
oddsratio(tab, method = "wald")
```

```{r}
pm <- data.frame(
before = c("A", "A", "B", "B"),
after = c("A", "B", "A", "B"),
count = c(71, 31, 14, 84)
)
# Turn into a contingency table with weights
tab <- xtabs(count ~ before + after, data = pm)
tab
# Agreement statistics
mcnemar.test(tab,correct=FALSE)
```

```{r}
nuggets <- read.csv('C:/Users/tyler/OneDrive/Documents/GitHub/work_to_show/Classes Spring 2026/BDA Spring 26/Datasets/nuggets.csv', sep = ',')
nuggets

head(nuggets)
model <- lm(temp ~ weight, data = nuggets)
summary(model)
#below this line “extra” code to provide a visual plot
plot(nuggets$weight,nuggets$temp)
abline(model)
pred_interval <- predict(model, newdata=nuggets, interval="prediction",level = 0.95)
lines(nuggets$weight, pred_interval[,2])
lines(nuggets$weight, pred_interval[,3])

```

```{r}
tendersmr <- read.csv('C:/Users/tyler/OneDrive/Documents/GitHub/work_to_show/Classes Spring 2026/BDA Spring 26/Datasets/tendersmr.csv', sep = ',')

head(tendersmr)
model <- lm(farmscore ~ Gender+live+tenders , data = tendersmr)
#alternate syntax to include all other columns:
# model <- lm(farmscore ~ ., data = tendersmr)
summary(model)
```

```{r}
nuggets <- read.csv('C:/Users/tyler/OneDrive/Documents/GitHub/work_to_show/Classes Spring 2026/BDA Spring 26/Datasets/nuggets.csv', sep = ',')
nuggets

head(nuggets)
model <- lm(temp ~ weight, data = nuggets)
summary(model)
#below this line “extra” code to provide a visual plot
plot(nuggets$weight,nuggets$temp)
abline(model)
pred_interval <- predict(model, newdata=nuggets, interval="prediction",level = 0.95)
lines(nuggets$weight, pred_interval[,2])
lines(nuggets$weight, pred_interval[,3])

```

```{r}
tendersmr <- read.csv('C:/Users/tyler/OneDrive/Documents/GitHub/work_to_show/Classes Spring 2026/BDA Spring 26/Datasets/tendersmr.csv', sep = ',')

head(tendersmr)
model <- lm(farmscore ~ Gender+live+tenders , data = tendersmr)
#alternate syntax to include all other columns:
# model <- lm(farmscore ~ ., data = tendersmr)
summary(model)
```

```{r}
nuggets <- read.csv('C:/Users/tyler/OneDrive/Documents/GitHub/work_to_show/Classes Spring 2026/BDA Spring 26/Datasets/nuggets.csv', sep = ',')
nuggets

head(nuggets)
model <- lm(temp ~ weight, data = nuggets)
summary(model)
#below this line “extra” code to provide a visual plot
plot(nuggets$weight,nuggets$temp)
abline(model)
pred_interval <- predict(model, newdata=nuggets, interval="prediction",level = 0.95)
lines(nuggets$weight, pred_interval[,2])
lines(nuggets$weight, pred_interval[,3])

```

```{r}
tendersmr <- read.csv('C:/Users/tyler/OneDrive/Documents/GitHub/work_to_show/Classes Spring 2026/BDA Spring 26/Datasets/tendersmr.csv', sep = ',')

head(tendersmr)
model <- lm(farmscore ~ Gender+live+tenders , data = tendersmr)
#alternate syntax to include all other columns:
# model <- lm(farmscore ~ ., data = tendersmr)
summary(model)
```

# R code for Logistic Regression
```{r}
p <- data.frame(
method = c("A", "A", "B", "B"),
quality = c("Good", "Defective", "Good", "Defective"),
count = c(75, 25, 87, 13)
)
# Convert quality to a factor, with "Defective" as the reference
p$quality <- factor(p$quality, levels = c("Defective", "Good"))
model <- glm(quality ~ method,
data = p,
family = binomial,
weights = count)
#global model test:
null_model<-glm(quality~1, family=binomial, data=p,
weights = count)
anova(null_model,model,test="Chisq")
summary(model)
exp(cbind(OR = coef(model), confint(model)))
```
R Code for Multiple Logistic Regression
```{r}
head(ps)
model <- glm(Purchase~Configuration+Size,
             data = ps,
             family = binomial)
summary(model)
null_model<-glm(Purchase~1, family=binomial, data=ps)
anova(null_model,model,test="Chisq")
summary(model)
exp(cbind(OR = coef(model), confint(model)))
levels(ps$Purchase)
```
Changing of ref level
```{r}
ps$Configuration <- relevel(ps$Configuration, ref = "C")
model <- glm(Purchase~Configuration+Size,
data = ps,
family = binomial)
summary(model)
null_model<-glm(Purchase~1, family=binomial, data=ps)
anova(null_model,model,test="Chisq")
summary(model)
exp(cbind(OR = coef(model), confint(model)))
levels(ps$Purchase)
```
R code for logistic regression
```{r}
actbda <-read.csv("C:/Users/tyler/OneDrive/Documents/GitHub/work_to_show/Classes Spring 2026/BDA Spring 26/Datasets/actbda.csv", header=TRUE)
head(actbda)
actbda$BDAsuccess<-factor(actbda$BDAsuccess)
model <- glm(BDAsuccess~ ACT,
data = actbda,
family = binomial)
null_model<-glm(BDAsuccess~1, family=binomial,
data=actbda)
anova(null_model,model,test="Chisq")
summary(model)
exp(cbind(OR = coef(model), confint(model)))
```
Regression for prediction
```{r}
# R Code to set up the cross-validation
house <-read.csv("C:/Users/tyler/OneDrive/Documents/GitHub/work_to_show/Classes Spring 2026/BDA Spring 26/Datasets/housing.csv", header=TRUE)
head(house)
colnames(house) # "sqft"      "saleprice"
# install.packages("caret") #only needs to be installed once
library(caret)
 
#we set a seed so that the random sampling will be repeatable
set.seed(42)
 
# Set up 5-fold cross-validation
train_control <- trainControl(method = "cv",number = 5)
 
# fit the model
model <- train(saleprice ~sqft, data = house,
               method = "lm",
               trControl = train_control)

# Evaluating the model
print(model)

# Linear Regression 
# 
# 60 samples
#  1 predictor
# 
# No pre-processing
# Resampling: Cross-Validated (5 fold) 
# Summary of sample sizes: 48, 48, 48, 48, 48 
# Resampling results:
# 
#   RMSE          # Mult RMSE x2 to get +- how off it is. 
#   8162.942  

# Rsquared
# 0.960561  

# MAE # MAE how on average you are wrong. 
# 6015.2

summary(model)

# Call:
# lm(formula = .outcome ~ ., data = dat)
# 
# Residuals: Residual information that we will ignore because we will look at a plot that will tell us more than this
#    Min     1Q Median     3Q    Max 
# -20448  -3536  -2110   2285  21115 
# 
# Coefficients: These are the coefficients (or slope and intercept) of our regression model The values to the right with *** next to it, signify that the slope and intercept are significant

#               Estimate Std. Error t value
# (Intercept) -31246.873   3485.634  -8.964
# sqft            78.812      2.221  35.483
#             Pr(>|t|)    
# (Intercept) 1.52e-12 ***
# sqft         < 2e-16 ***
# ---
# Signif. codes:  
# 0 ‘***’ 0.001 ‘**’ 0.01 ‘*’ 0.05 ‘.’ 0.1 ‘ ’ 1
# 
# Residual standard error: 8030 on 58 degrees of freedom
# Multiple R-squared:  0.956,	Adjusted R-squared:  0.9552 
# F-statistic:  1259 on 1 and 58 DF,  p-value: < 2.2e-16

# ================================================
# ***The p-value of the F-statistics also tells us the regression model is significant***
# ================================================

regmodel<-lm(saleprice~sqft,data=house)
plot(house$sqft,house$saleprice)
abline(regmodel)
pred_interval <- predict(regmodel, newdata=house, interval="prediction",level = 0.95)
lines(house$sqft, pred_interval[,2])
lines(house$sqft, pred_interval[,3])

# make a prediction
newdata=data.frame(sqft=
                     1400 # make the prediction within the range 
                   )
predict(regmodel,newdata=newdata,interval="prediction",level=0.95) 

# ================================================
# This code gives the predicted sale price for a square footage of 1400
# It gives our best estimate for the sale price and then a lower and upper bound for two sale prices we are 95% sure the true sale price will be between.
# 
# What sqft values is the model valid for?
# Use the following to see the min and max values of sqft:
# 
# summary(house)
# 
# The model can be used for values between 620 and 3410 square feet.
# ================================================
```
```{r}
df <- read.csv("C:/Users/tyler/OneDrive/Documents/GitHub/work_to_show/Classes Spring 2026/BDA Spring 26/Datasets/tenders.csv")

head(df)
colnames(df) # "live"    "tenders"
# install.packages("caret") #only needs to be installed once
library(caret)
 
#we set a seed so that the random sampling will be repeatable
set.seed(42)
 
# Set up 5-fold cross-validation
train_control <- trainControl(method = "cv",number = 5)
 
# fit the model
model <- train(tenders ~live, data = df,
               method = "lm",
               trControl = train_control)

# Evaluating the model
print(model)

# Linear Regression 
# 
# 77 samples
#  1 predictor
# 
# No pre-processing
# Resampling: Cross-Validated (5 fold) 
# Summary of sample sizes: 63, 61, 61, 62, 61 
# Resampling results:
# 
#   RMSE      Rsquared   MAE     
#   9.198435  0.8049075  7.386984
# 
# Tuning parameter 'intercept' was held constant
#  at a value of TRUE

# summary(model)
# 
# Call:
# lm(formula = .outcome ~ ., data = dat)
# 
# Residuals:
#      Min       1Q   Median       3Q      Max 
# -19.7853  -4.7827   0.2867   6.1831  20.5290 
# 
# Coefficients:
#              Estimate Std. Error t value Pr(>|t|)    
# (Intercept) 14.519579   6.639455   2.187   0.0319 *  
# live         0.035215   0.002162  16.288   <2e-16 ***
# ---
# Signif. codes:  0 ‘***’ 0.001 ‘**’ 0.01 ‘*’ 0.05 ‘.’ 0.1 ‘ ’ 1
# 
# Residual standard error: 9.148 on 75 degrees of freedom
# Multiple R-squared:  0.7796,	Adjusted R-squared:  0.7767 
# F-statistic: 265.3 on 1 and 75 DF,  p-value: < 2.2e-16

# ================================================
# ***The p-value of the F-statistics also tells us the regression model is significant***
# ================================================

regmodel<-lm(tenders~live,data=df)
plot(df$live,df$tenders)
abline(regmodel)
pred_interval <- predict(regmodel, newdata=df, interval="prediction",level = 0.95)
lines(df$live, pred_interval[,2])
lines(df$live, pred_interval[,3])

summary(df) # see min and max
# make a prediction
newdata=data.frame(live=
                     2260 # make the prediction within the range 
                   )
predict(regmodel,newdata=newdata,interval="prediction",level=0.95) 

# ================================================
# This code gives the predicted sale price for a square footage of 1400
# It gives our best estimate for the sale price and then a lower and upper bound for two sale prices we are 95% sure the true sale price will be between.
# 
# What live values is the model valid for?
# Use the following to see the min and max values of live:
# 

# summary(df)

# 
# The model can be used for values between 620 and 3410 square feet.
# ================================================
```

```{r}
# read in program
house2 <- read.csv("C:/Users/tyler/OneDrive/Documents/GitHub/work_to_show/Classes Spring 2026/BDA Spring 26/Datasets/housing2.csv")
head(house2)
colnames(house2)
library(caret)
set.seed(42)
train_control <- trainControl(method = "cv",number = 5)
model <- train(SalePrice ~., data = house2, method = "lm", trControl =
train_control)
print(model)
summary(model)
regmodel<-lm(SalePrice~.,data=house2)
summary(house2) #to see max and min of inputs
newdata=data.frame(SqFt=1400,Age=4,Lotsize=.5)
predict(regmodel,newdata=newdata,interval="prediction",level=0.95)
```
R for Logistic Regression
```{r}
# R for Logistic Regression
# example 1
stu <-read.csv("C:/Users/tyler/OneDrive/Documents/GitHub/work_to_show/Classes Spring 2026/BDA Spring 26/Datasets/student success (2).csv", header=TRUE)
head(stu)
stu$success<-factor(stu$success)
set.seed(23)
R=runif(nrow(stu)) #create a random uniform column the length of our table
stu$R=R #add the random numbers to our table
train<-stu[stu$R<=.6,] #select roughly 60% for training
test<-stu[stu$R>.6,] #the other 40% gets placed in test
train<-subset(train,select=-c(R)) #remove the column R
test<-subset(test,select=-c(R)) #remove the column R

# log regression
lreg<-glm(success~FinalGrade1403+Transfer+ACT_Score+ACT_Math+HS_GPA, data=train, family=binomial)
summary(lreg)

# predict model
pmodel<- predict(lreg,newdata=test, type = "response")

model<- ifelse(pmodel > 0.5, "yes", "no") #turn the probability into a pred
known<-test["success"]
knownt<-t(known)
table(knownt,model)
newdata=data.frame(FinalGrade1403=87,Transfer="yes",ACT_Score=26,ACT_Math=25,HS_GPA=3.9)
predict(lreg, newdata = newdata, type = "response") 
ifelse(predict(lreg, newdata = newdata, type = "response")>.5, "yes", "no")
```
```{r}
# R for Logistic Regression
# example 2
appt <-read.csv("C:/Users/tyler/OneDrive/Documents/GitHub/work_to_show/Classes Spring 2026/BDA Spring 26/Datasets/appt (1).csv", header=TRUE)
head(appt) # SHOW DATA 
appt$MissedAppt<-factor(appt$MissedAppt) # TREATE AS FACTOR
set.seed(23) # SET THE SEED
R=runif(nrow(appt)) #create a random uniform column the length of our table
appt$R=R #add the random numbers to our table
train<-appt[appt$R<=.6,] #the other 40% gets placed in test
test<-appt[appt$R>.6,] #the other 40% gets placed in test
train<-subset(train,select=-c(R)) #remove the column R
test<-subset(test,select=-c(R)) #remove the column R
lreg<-glm(MissedAppt~., data=train, family=binomial)
summary(lreg)
pmodel<- predict(lreg,newdata=test, type = "response")
model<- ifelse(pmodel > 0.5, "yes", "no") 
known<-test["MissedAppt"]
knownt<-t(known)
table(knownt,model)
library(pROC)
test_prob = predict(lreg, newdata = test, type = "response")
test_roc = roc(test$MissedAppt ~ test_prob, plot = TRUE, print.auc = TRUE)

# ACCURACY OF THE MODEL
ACCURACY = (156 + 535) / (156 + 154 + 51 + 535)
ACCURACY

# MAKE THE PREDICITON 
predict_data=data.frame(PrevMissedAppt="no",
                   SameDayAppt="yes", 
                   Age=47, Male="yes", 
                   White="yes", 
                   WaitTime=30, 
                   Monday="yes", 
                   Tuesday="no", 
                   Thursday="no", 
                   Friday="no", 
                   Saturday="no"
                  )
predict(lreg, # logistic regression
        predict_data, # prediction dataframe 
        type = "response") # greater than .5 they are going to miss 

#Since this is greater than 0.5, the prediction is “yes” the appointment will be missed
ifelse(predict(lreg,
               newdata = predict_data, 
               type = "response")
       >.5, 
       "yes", # >.5 going to miss 
       "no" # <.5 not going to miss 
      )
```
# CLASSIFICATION MODELS

LOGISIC REGRESSION
```{r}
target <- "Survived"
df <-read.csv("C:/Users/tyler/OneDrive/Documents/GitHub/work_to_show/Classes Spring 2026/BDA Spring 26/Datasets/titanic.csv", header=TRUE)
head(df)
colnames(df) # 'Class''Age''Gender''Survived'
df$Survived<-factor(df$Survived) # make success a factor
# -----------------------------------------------------------------------------
# divide into train and test data
set.seed(23)
R=runif(nrow(df)) #create a random uniform column the length of our table
df$R=R #add the random numbers to our table
train<-df[df$R<=.6,] #select roughly 60% for training
test<-df[df$R>.6,] #the other 40% gets placed in test
train<-subset(train,select=-c(R)) #remove the column R
test<-subset(test,select=-c(R)) #remove the column R

# MAKE A PREDICTION 

# 'Class''Age''Gender''Survived'

predict_data=data.frame(

Class=3, 

Age=18,

Gender="F"

)

lreg<-glm(Survived~., data=train, family=binomial)

summary(lreg)

pmodel<- predict(lreg,newdata=test, type = "response")

model<- ifelse(pmodel > 0.5, "yes", "no") 

known<-test["Survived"]

knownt<-t(known)

table(knownt,model)

library(pROC)

test_prob = predict(lreg, newdata = test, type = "response")

test_roc = roc(test$Survived ~ test_prob, plot = TRUE, print.auc = TRUE)

# =====================================================================

# ACCURACY OF THE MODEL

n_n <- 160

n_y <- 41

y_n <-31

y_y <- 88

ACCURACY <- (n_n + y_y) / (n_n + n_y + y_n + y_y)

ACCURACY

# =====================================================================

# unique numbers to help with prediction

unique(df$Class)

unique(df$Age)

unique(df$Gender)

# =====================================================================

# MAKE A PREDICTION 

# 'Class''Age''Gender''Survived'

predict(lreg, # logistic regression

predict_data, # prediction dataframe 

type = "response") # greater than .5 they are going to miss 

# =====================================================================

# if else statement to help put a name to the number

ifelse(predict(lreg,

newdata = predict_data, 

type = "response")

>.5, 

"survived", # >.5 going to survive 

"did not survive" # <.5 not going to survive

)

# c1
no_no <- 160
no_yes <- 41
# c2
yes_no <- 31
yes_yes <-88

accuracy_lr <- (no_no + yes_yes) / (no_no + no_yes + yes_no + yes_yes)
round(accuracy_lr,2)
error_rate_lr <- (yes_no + no_yes) / (no_no + no_yes + yes_no + yes_yes)
round(error_rate_lr,2)
true_no_rate_lr <- no_no / (no_no + no_yes)
round(true_no_rate_lr,2)
true_yes_rate_lr <- yes_yes / (yes_no + yes_yes)
round(true_yes_rate_lr,2)
```

CLASS TREE
```{r}
target <- "Survived"
df <-read.csv("C:/Users/tyler/OneDrive/Documents/GitHub/work_to_show/Classes Spring 2026/BDA Spring 26/Datasets/titanic.csv", header=TRUE)
head(df)
colnames(df) # 'Class''Age''Gender''Survived'
df$Survived<-factor(df$Survived) # make success a factor
# -----------------------------------------------------------------------------
# divide into train and test data
set.seed(23)
R=runif(nrow(df)) #create a random uniform column the length of our table
df$R=R #add the random numbers to our table
train<-df[df$R<=.6,] #select roughly 60% for training
test<-df[df$R>.6,] #the other 40% gets placed in test
train<-subset(train,select=-c(R)) #remove the column R
test<-subset(test,select=-c(R)) #remove the column R

# MAKE A PREDICTION 

# 'Class''Age''Gender''Survived'

predict_data=data.frame(

Class=3, 

Age=18,

Gender="F"

)

# class tree
#install.packages('rpart', dependencies = TRUE)
#install.packages('rpart.plot', dependencies = TRUE)
library(rpart)
library(rpart.plot)
#can’t use success~. due to extra columns in train:
ctree<-rpart(Survived~.,data=train)
#summary(ctree)
#plot the tree, type=4 labels splits and the nodes, extra=101 adds extra calculations in each node
rpart.plot(ctree, type = 4, extra = 101)

# fit the model to the testing data
model<- predict(ctree, newdata=test, type = "class")
known<-test[target] #extract the actual value of success for comparing
knownt<-t(known) #transpose the data for the confusion matrix
table(knownt,model) #create the confusion matrix

# interpreting the output CLASS TREE
# c1
no_no <- 166
no_yes <- 43
# c2
yes_no <- 25
yes_yes <-86

accuracy_tree <- (no_no + yes_yes) / (no_no + no_yes + yes_no + yes_yes)
round(accuracy_tree,2)
error_rate_tree <- (yes_no + no_yes) / (no_no + no_yes + yes_no + yes_yes)
round(error_rate_tree,2)
true_no_rate_tree <- no_no / (no_no + no_yes)
round(true_no_rate_tree,2)
true_yes_rate_tree <- yes_yes / (yes_no + yes_yes)
round(true_yes_rate_tree,2)

#predict with the classification tree
predict(ctree,newdata=predict_data,type="class")
```
RANDOM FORREST
```{r}
target <- "Survived"
df <-read.csv("C:/Users/tyler/OneDrive/Documents/GitHub/work_to_show/Classes Spring 2026/BDA Spring 26/Datasets/titanic.csv", header=TRUE)
head(df)
colnames(df) # 'Class''Age''Gender''Survived'
df$Survived<-factor(df$Survived) # make success a factor
# -----------------------------------------------------------------------------
# divide into train and test data
set.seed(23)
R=runif(nrow(df)) #create a random uniform column the length of our table
df$R=R #add the random numbers to our table
train<-df[df$R<=.6,] #select roughly 60% for training
test<-df[df$R>.6,] #the other 40% gets placed in test
train<-subset(train,select=-c(R)) #remove the column R
test<-subset(test,select=-c(R)) #remove the column R

# MAKE A PREDICTION 

# 'Class''Age''Gender''Survived'

predict_data=data.frame(

Class=3, 

Age=18,

Gender="F"

)

# random forrest
library(randomForest)
set.seed(23) #needed to make the RF repeatable
RFM=randomForest(Survived~.,data=train)
#summary(RFM)

# fit model to testing data
model<- predict(RFM, newdata=test, type = "class")
known<-test[target] #extract the actual value of success for comparing
knownt<-t(known) #transpose the data for the confusion matrix
table(knownt,model) #create the confusion matrix

# fine tune the model
set.seed(23)
RFM=randomForest(Survived~.,data=train,
mtry=4,ntrees=1000)
model<- predict(RFM, newdata=test, type = "class")
known<-test[target] #extract the actual value of success for comparing
knownt<-t(known) #transpose the data for the confusion matrix
table(knownt,model) #create the confusion matrix

# interpreting the output random forrest
# c1
no_no <- 151
no_yes <- 36
# c2
yes_no <- 40
yes_yes <- 93

accuracy_rf <- (no_no + yes_yes) / (no_no + no_yes + yes_no + yes_yes)
round(accuracy_rf,2)
error_rate_rf <- (yes_no + no_yes) / (no_no + no_yes + yes_no + yes_yes)
round(error_rate_rf,2)
true_no_rate_rf <- no_no / (no_no + no_yes)
round(true_no_rate_rf,2)
true_yes_rate_rf <- yes_yes / (yes_no + yes_yes)
round(true_yes_rate_rf,2)

# predict w/ rf
predict(RFM, newdata = predict_data, type = "class")
```
ANN
```{r}
target <- "Survived"
df <-read.csv("C:/Users/tyler/OneDrive/Documents/GitHub/work_to_show/Classes Spring 2026/BDA Spring 26/Datasets/titanic.csv", header=TRUE)
head(df)
colnames(df) # 'Class''Age''Gender''Survived'
df$Survived<-factor(df$Survived) # make success a factor
# -----------------------------------------------------------------------------
# divide into train and test data
set.seed(23)
R=runif(nrow(df)) #create a random uniform column the length of our table
df$R=R #add the random numbers to our table
train<-df[df$R<=.6,] #select roughly 60% for training
test<-df[df$R>.6,] #the other 40% gets placed in test
train<-subset(train,select=-c(R)) #remove the column R
test<-subset(test,select=-c(R)) #remove the column R

# MAKE A PREDICTION 

# 'Class''Age''Gender''Survived'

predict_data=data.frame(

Class=3, 

Age=18,

Gender="F"

)

# ann
#install.packages("nnet")
library(nnet)
set.seed(17)
ANN<-nnet(Survived~.,data=train, size=4)

model<- predict(ANN, newdata=test, type = "class")
known<-test[target] #extract the actual value of success for comparing
knownt<-t(known) #transpose the data for the confusion matrix
table(knownt,model) #create the confusion matrix

# interpret the model
# c1
no_no <- 169
no_yes <- 42
# c2
yes_no <- 22
yes_yes <- 87

accuracy_ann <- (no_no + yes_yes) / (no_no + no_yes + yes_no + yes_yes)
round(accuracy_ann,2)
error_rate_ann <- (yes_no + no_yes) / (no_no + no_yes + yes_no + yes_yes)
round(error_rate_ann,2)
true_no_rate_ann <- no_no / (no_no + no_yes)
round(true_no_rate_ann,2)
true_yes_rate_ann <- yes_yes / (yes_no + yes_yes)
round(true_yes_rate_ann,2)

# ann prediction
predict(ANN,newdata=predict_data, type="class")
```

ACCURACY OF CLASS MODELS
```{r}
cat("ACCURACY: ", "rf ", round(accuracy_rf,2), 
    " tree ", round(accuracy_tree,2),
    " ann ", round(accuracy_ann,2),
    "lr ", round(accuracy_lr,2))

accuracies <- c(
    "Random Forest" = accuracy_rf,
    "Decision Tree" = accuracy_tree,
    "ANN" = accuracy_ann,
    "Log Regression" = accuracy_lr
)

best_model <- names(which.max(accuracies))
print(best_model)
```

# VAR SELECTION

Example 1
The file “product survey.csv” are data from 40 different products from the same product category for a large online retail company. The column “sales” is the average weekly sales of each item. Columns, “color”, “size”, “features”, “reliability”, “warranty”, and “cost” are the average scores from product surveys that ask about the customer’s satisfaction with each aspect of the product. A score closer to 10 indicates a higher average customer satisfaction for each item. Determine the best subset of variables to predict sales.
```{r}
# Stepwise Regression using AIC
ps <-read.csv("C:/Users/tyler/OneDrive/Documents/GitHub/work_to_show/Classes Spring 2026/BDA Spring 26/Datasets/product survey (3).csv", header=TRUE)
head(ps)
# install.packages("MASS")
library(MASS)
step_model <- stepAIC(lm(Sales ~ ., data = ps), direction = "both", trace = FALSE)
# Print the selected variables
print(step_model$anova)

cat(
"  *****The final variables selected were: Size, Features, Reliability, and Cost*****
  *****Color and Warranty were not selected*****"
)

# Random Forest
library(randomForest)
ps$Sales<-factor(ps$Sales)
RFM=randomForest(Sales~.,data=ps, importance=TRUE)
importance(RFM)
varImpPlot(RFM)

cat(
"  Importance using MSE is given in the first graph; importance using Node Purity is given in the second—either could be used to determine which variables are most useful for building a model

One of the drawbacks to using Random Forests for variable selection is that, while variables are ranked by importance, there is no clear “cut-off” for where to stop including variables (it is left up to analyst to say where to stop including variables)

However, in this case, warrant and color are less than 0 on %IncMSE, clearly indicating they should not be used."
)

# Lasso Regression
# Note these variables are all on the same scale, so no standardization is needed.
# install.packages("gamlr")
library(gamlr)
modMat<-model.matrix(~Color+Size+Features+Reliability+Warranty+Cost, data=ps)
fit<-gamlr(x=modMat,y=ps$Sales)
plot(fit)
coef(fit)

cat(
  "8 x 1 sparse Matrix of class dgCMatrix
                seg67
intercept   3338.6351
(Intercept)    .     <- Dot indicates not selected
Color          .     <- Dot indicates not selected
Size         302.2088
Features     365.7851
Reliability  382.6246
Warranty       .     <- Dot indicates not selected
Cost        1041.1788

The regularization path image can be nice to see,
however it is the output above that tells us Color
and Warranty should not be selected. (Both have
dots instead of parameter estimates)"
)

cat(
"  # Summary example 1
• All three methods of variable selection agreed that Color and
Warranty should not be used in the model

• The process for building the final model would now begin using just
the selected columns: Size, Features, Reliability, and Cost"
)
```
Example 2
The file “irs2.csv” contains data from past US Internal Revenue Service audits. In an effort to more efficiently audit only those returns most likely to have violated tax laws, the columns in the file irs2.csv are:

• cheated: "yes" if the audit found the return violated tax laws, otherwise "no" -- this is the target
variable we are building the model to predict

• married: "yes" if the return is for a married individual; "no" for single

• dependents: "yes" if the return is for an individual claiming dependents; "no" otherwise

• incomeOver100K: "yes" if the income of the individual exceeds $100,000

• return: "yes" if the individual received a return ("no" if a the individual did not receive a return or
had to pay in)

• Largedonation: “yes” if the individual made a donation in excess of 10% of their annual income,
no otherwise

• mathErrors: “yes”: if there were math errors in the return, “no” otherwise

• filedLate: "yes" if the return was filed late, "no" if filed early or on time

• Over4DeductClaimed: "yes" if over 4 deductions were claimed, "no" if 4 or less

```{r}
irs <-read.csv("C:/Users/tyler/OneDrive/Documents/GitHub/work_to_show/Classes Spring 2026/BDA Spring 26/Datasets/irs2.csv", header=TRUE)
# head(irs)
# ==============================================
# An extra step is required for stepwise selection for categorical data, to turn the “yes” and “no”s into 1’s and 0’s (called dummy variables or indicator variables)
# ==============================================
# Dummy code categorical predictor variables
irsdum <- model.matrix(~., irs)[,-1]
irsdum<-data.frame(irsdum)
head(irsdum)
#install.packages("MASS")
library(MASS)
step_model <- stepAIC(lm(cheatedyes ~ ., data = irsdum, family=binomial), direction = "both", trace = FALSE)
# Print the selected variables
print(step_model$anova)

cat(
  "Stepwise Model Path 
Analysis of Deviance Table

Initial Model:
cheatedyes ~ marriedyes + dependentsyes + incomeOver100Kyes + 
    returnyes + largedonationyes + mathErrorsyes + filedLateyes + 
    Over4DeductClaimedyes

Final Model:
cheatedyes ~ marriedyes + incomeOver100Kyes + returnyes + filedLateyes + 
    Over4DeductClaimedyes


                Step Df    Deviance Resid. Df Resid. Dev       AIC
1                                         477   41.48767 -1177.955
2    - dependentsyes  1 0.006368431       478   41.49404 -1179.880
3    - mathErrorsyes  1 0.010191336       479   41.50423 -1181.761
4 - largedonationyes  1 0.078835478       480   41.58306 -1182.839

• The variables married, incomeOver100, return, filedLate, and Over4DeductClaim were selected. (Depedents, mathErrors, and largedonation were not selected)"
)

# Random Forest
# head(irs)
irs$cheated<-factor(irs$cheated)
RFM=randomForest(cheated~.,data=irs, importance=TRUE)
importance(RFM)
varImpPlot(RFM)
# (Note, there is no need to recode the data into dummy variables for random forests)
cat("Importance using accuracy is given in the first graph; importance using the Gini criterion is given in the second—either could be used to determine which variables are most useful for building a model

In this case, there is a clear drop-off for largedonation, mathErrors, and dependents, indicating they should not be used.")

# Dummy code categorical predictor variables
irsdum <- model.matrix(~., irs)[,-1]
irsdum<-data.frame(irsdum)
#head(irsdum)
#lasso regression
#install.packages("gamlr")
library(gamlr)
modMat<-model.matrix(~marriedyes + dependentsyes + incomeOver100Kyes +
returnyes + largedonationyes + mathErrorsyes + filedLateyes +
Over4DeductClaimedyes, data=irsdum)
fit<-gamlr(x=modMat,y=irsdum$cheatedyes, family = "binomial")
plot(fit)
coef(fit)

# Note these variables are all on the same scale, so no standardization is needed.
# However, dummy variables will again need to be created for the lasso technique

cat("
    10 x 1 sparse Matrix of class dgCMatrix
                           seg99
intercept             -2.5741270
(Intercept)            .        
marriedyes            -3.6222103
dependentsyes          .        
incomeOver100Kyes      4.1429629
returnyes             -4.7533481
largedonationyes      -0.1616757
mathErrorsyes          .        
filedLateyes           4.2112906
Over4DeductClaimedyes  4.0355239

Only dependents and mathErrors were removed; however, largedonation departed from zero late in the process and may not be needed; in this case both stepwise and random forests can help clarify whether to include largedonations as a predictor

# Summary example 2
• All three methods of variable selection agreed that dependents and
mathErrors should not be used in the model
• Stepwise and Random Forests indicated largedonation should also be
removed
• Further, largedonation was selected late in the regularization path
from the Lasso
• The final model selected would likely not use these three columns
and instead be based on married, incomeOver100K, return, filedLate,
and Over4DeductClaimedyes (these were the variables you used to
build your ANN in week 8)
")
```
Homework
```{r}
nfl <- read.csv("C:/Users/tyler/OneDrive/Documents/GitHub/work_to_show/Classes Spring 2026/BDA Spring 26/Datasets/nflcombine2021.csv")

# step wise
#install.packages("MASS")
library(MASS)
step_model <- stepAIC(lm(Draftedyes ~ ., data = nfl, family=binomial), direction = "both", trace = FALSE)
# Print the selected variables
print(step_model$anova)

# random forest
library(randomForest)
nfl$Draftedyes <-factor(nfl$Draftedyes)
RFM=randomForest(Draftedyes ~.,data=nfl, importance=TRUE)
importance(RFM)
varImpPlot(RFM)
           
znfl <- read.csv("C:/Users/tyler/OneDrive/Documents/GitHub/work_to_show/Classes Spring 2026/BDA Spring 26/Datasets/znflcombine2021.csv")

#lasso regression
#install.packages("gamlr")
library(gamlr)
modMat<-model.matrix(~ Ht + Wt + yd40 + Vertical + Bench + BroadJump +
Cone3 + Shuttle, data=znfl)
fit<-gamlr(x=modMat,y=znfl$Draftedyes, family = "binomial")
#the plot shows the coeff moving from zero to values
plot(fit)
#selected model
coef(fit)

cat(
  "
  All methods agree Wt and yd40 are the most important variables to predict if a player is likely to be drafted—these
should definitely be included in a final model. Beyond that, there is no clear consensus and several factors could be
included or left out. In this case, I would probably side with LASSO and use all factors.
  ")
```

# association analysis
```{r}
survey <- read.csv("C:/Users/tyler/OneDrive/Documents/GitHub/work_to_show/Classes Spring 2026/BDA Spring 26/Datasets/survey.csv")
head(survey)
colnames(survey)

# install.packages("arules") #run only once
library("arules")

#identify transactions
transactions <- as(split(survey$Answer, survey$Stdnt_Nbr),"transactions") # specifies how data will be read in

# use the apriori algorithm for association analysis
# specify the minimum support and minimum confidence
# maxlen=2 tells R to only consider pairwise associations
rules<-apriori(transactions,
               parameter=
                 list(support=0.02,confidence=0.1,maxlen=2) # thresholds
)

#include the odds ratio, use the Haldane-Anscombe correction, sort by odds ratio
ci <- confint(rules, "oddsRatio", smoothCounts = .5)

quality(rules) <- cbind(
  quality(rules),
  oddsRatio = interestMeasure(rules, "oddsRatio", smoothCounts = .5),
  oddsRatio = ci)

rules <- sort(rules, by = "oddsRatio")
inspect(rules)

# Convert rules to a data frame and export to CSV
rules_df <- as(rules, "data.frame")

# Also add LHS and RHS as separate columns for readability
rules_df$lhs <- labels(lhs(rules))
rules_df$rhs <- labels(rhs(rules))

write.csv(rules_df, "C:/Users/tyler/OneDrive/Documents/GitHub/work_to_show/Classes Spring 2026/BDA Spring 26/survey_rules.csv", row.names = FALSE)

cat("Exported", nrow(rules_df), "rules to CSV.\n")
```

```{r}
music <- read.csv("C:/Users/tyler/OneDrive/Documents/GitHub/work_to_show/Classes Spring 2026/BDA Spring 26/Datasets/music.csv")
colnames(music)
head(music)

# Customers who added “TheShins” to their favorites list were 3.8 times more likely to add “VampireWeekend” to their favorites list.
#install.packages("arules") #run only once
library("arules")

#identify transactions
transactions <- as(split(music$artist, music$CustomerID),"transactions")

#use the apriori algorithm for association analysis
#specify the minimum support and minimum confidence
#maxlen=2 tells R to only consider pairwise associations
rules<-apriori(transactions,parameter=list(support=0.02,confidence=0.1,maxlen=2))

#include the odds ratio, use the Haldane-Anscombe correction, sort by odds ratio
ci <- confint(rules, "oddsRatio", smoothCounts = .5)
quality(rules) <- cbind(
quality(rules),
oddsRatio = interestMeasure(rules, "oddsRatio", smoothCounts = .5),
oddsRatio = ci)
rules <- sort(rules, by = "oddsRatio")
inspect(rules)

# Convert rules to a data frame and export to CSV
rules_df <- as(rules, "data.frame")

# Also add LHS and RHS as separate columns for readability
rules_df$lhs <- labels(lhs(rules))
rules_df$rhs <- labels(rhs(rules))

write.csv(rules_df, "C:/Users/tyler/OneDrive/Documents/GitHub/work_to_show/Classes Spring 2026/BDA Spring 26/music_rules.csv", row.names = FALSE)

cat("Exported", nrow(rules_df), "rules to CSV.\n")
```

homework
```{r}
df <- read.csv("C:/Users/tyler/OneDrive/Documents/GitHub/work_to_show/Classes Spring 2026/BDA Spring 26/Datasets/mb.csv")
colnames(df)
head(df)

#install.packages("arules") #run only once
library("arules")
#identify transactions
# transactions <- as(split(df$y, df$x),"transactions")
transactions <- as(split(df$itemtype, df$visitid),"transactions")

#use the apriori algorithm for association analysis
#specify the minimum support and minimum confidence
#maxlen=2 tells R to only consider pairwise associations
rules<-apriori(transactions,
               parameter=
                 list(support=0.02,confidence=0.1,maxlen=2)
               )

#include the odds ratio, use the Haldane-Anscombe correction, sort by odds ratio
ci <- confint(rules, "oddsRatio", smoothCounts = .5)
quality(rules) <- cbind(
  quality(rules),
  oddsRatio = interestMeasure(rules, "oddsRatio", smoothCounts = .5),
  oddsRatio = ci)

rules <- sort(rules, by = "oddsRatio")
inspect(rules)

# Convert rules to a data frame and export to CSV
rules_df <- as(rules, "data.frame")

# Also add LHS and RHS as separate columns for readability
rules_df$lhs <- labels(lhs(rules))
rules_df$rhs <- labels(rhs(rules))

write.csv(rules_df, "C:/Users/tyler/OneDrive/Documents/GitHub/work_to_show/Classes Spring 2026/BDA Spring 26/mb_rules.csv", row.names = FALSE)

cat("Exported", nrow(rules_df), "rules to CSV.\n")
```

# P-VALUE REJECTION
```{r P-VALUE REJECTION, echo=TRUE, message=TRUE, warning=TRUE, paged.print=TRUE}
p_val <- 2e-16
level_sig <- 0.05

if (p_val <= level_sig) {
  print("Reject")
} else {
  print("Fail to reject")
}
```

