---
title: "EDA Assignment"
output: html_notebook
---


The file "student data for EDA.csv" contains exam, homework, and project scores for 119
students. 
-Using R, perform exploratory data analysis. Calculate summary statistics for the data
(min, max, mean, standard deviation). 
-Create histograms, run (line) charts, and scatter plots for each column (exam, homework, and project). 
-Clearly and specifically, identify any potential information quality issues in the data and provide the output the identified the issue. 
-Include your R program and the visualization or summary statistics that identified the issue.


```{r}
# EDA in R (full script)
df <-read.csv("C:/Users/tyler/OneDrive/Documents/GitHub/work_to_show/Classes Spring 2026/BDA Spring 26/student data for EDA.csv",header=TRUE) # FORWARD SLASH AND HEADER TRUE IF HAVE COLUMN NAMES
# column names
colnames(df) # "Student"  "Exam"     "Homework" "Project" 
#head of the data
head(df)
#summary stats
summary(df$Student)
summary(df$Exam)
summary(df$Homework)
summary(df$Project)
'''
> summary(df$Student)
   Min. 1st Qu.  Median    Mean 3rd Qu.    Max. 
    1.0    30.5    60.0    60.0    89.5   119.0 
> summary(df$Exam)
   Min. 1st Qu.  Median    Mean 3rd Qu.    Max. 
  30.82   66.14   79.39   75.65   90.19   99.40 
> summary(df$Homework)
   Min. 1st Qu.  Median    Mean 3rd Qu.    Max. 
 -77.73   73.03   88.34   80.80  100.00  100.00 
> summary(df$Project)
   Min. 1st Qu.  Median    Mean 3rd Qu.    Max. 
  29.56   73.34   87.18   82.17   99.17  100.00 
'''
# mean of columns
mean(df$Student)
mean(df$Exam)
mean(df$Homework)
mean(df$Project)
'''
> mean(df$Student)
[1] 60
> mean(df$Exam)
[1] 75.65008
> mean(df$Homework)
[1] 80.79866
> mean(df$Project)
[1] 82.16706
'''
# stand dev of columns
sd(df$Student)
sd(df$Exam)
sd(df$Homework)
sd(df$Project)
'''
> sd(df$Student)
[1] 34.49638
> sd(df$Exam)
[1] 18.20467
> sd(df$Homework)
[1] 23.82734
> sd(df$Project)
[1] 18.30348
'''
# finding the regression line 
mean(df$Student)+3*sd(df$Student)
mean(df$Student)-3*sd(df$Student)
mean(df$Exam)+3*sd(df$Exam)
mean(df$Exam)-3*sd(df$Exam)
mean(df$Homework)+3*sd(df$Homework)
mean(df$Homework)-3*sd(df$Homework)
mean(df$Project)+3*sd(df$Project)
mean(df$Project)-3*sd(df$Project)
'''
[1] 163.4891
[1] -43.48913
[1] 130.2641
[1] 21.03608
[1] 152.2807
[1] 9.316634
[1] 137.0775
[1] 27.25663
'''
#histogram
hist(df$Student)
hist(df$Exam)
hist(df$Homework) # some input is wrong
hist(df$Project)
#plot with ab line
plot(df$Student,type="b")
abline(h=mean(df$Student))

plot(df$Exam,type="b")
abline(h=mean(df$Exam))

plot(df$Homework,type="b") # extremely low outlier
abline(h=mean(df$Student))

plot(df$Project,type="b")
abline(h=mean(df$Exam))
#correlation
cor(df$Student, df$Exam)
plot(df$Student, df$Exam)

cor(df$Student, df$Homework) # outlier shows 
plot(df$Student, df$Homework)

cor(df$Student, df$Project)
plot(df$Student, df$Project)

cor(df$Exam, df$Homework) # outlier shows 
plot(df$Exam, df$Homework)

cor(df$Exam, df$Project)
plot(df$Exam, df$Project)

cor(df$Homework, df$Project) # outlier shows 
plot(df$Homework, df$Project)
```

```{r}
sub_0 <- sum(df$Homework < 0, na.rm = TRUE)
sub_0

```

